### 2023-03-15 Manuscript/Draft1.Rmd
updated manuscript and ratio plots

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error of abundance estimates and model output. General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors. We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covariates in VAST models actually increase the absolute error. When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.

@@

In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table. These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts. Also for Yellowtail, with all strata included adding covariates greatly improved estimates. With strata excluded there was still an improvement with covariates, but much less dramatic than with all strata.

Notes on differences between VAST settings: For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse). 80/96 ~ 83% of scenarios had a VAST run with a better error than the stratified mean. 15/16 of the times the stratified mean was better VAST was not using covariates. The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop_IncTemp_Allstrata_WCov_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP_IncTemP_allstrata for Had and YT). The overestimate in the fall only seems to be related to the way the population shifts between season. Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata). In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17). Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the model value, while the stratified mean estimates are, on average, too low. This can be further examined by viewing model/estimate ratio plots, in which VAST estimates remain much closer to the desired value of 1 compared to stratified mean estimates, which can range as high as 10 and exhibit large yearly changes.

A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 26\% of all VAST ratios were above 1 (84\% less than 1), with an average of 85\% with a standard deviation of 0.09. On the other hand, 67\% of stratified mean estimates were above 1 with an average of 1.51 and a standard deviation of 0.31. There were seasonal differences in estimate ratios with spring VAST ratios producing (0.96,0.07), fall VAST ratios being (0.74,0.14), stratified mean spring ratios resulting in (1.31,0.19), and fall values of (1.78,0.51). The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1. However, with a reduced domain adding covariates can decrease the accuracy of VAST estimates. When including all strata, the stratified mean produced (1.54,0.32), while adding covariates improved the VAST estimate ratio from (0.79,0.06) to (0.92,0.06). With a reduced number of strata, the stratified mean resulted in our worst summary values of (3.48,1.82), and adding covariates to our VAST models decreased the ratio results from (1.33,0.31 ) to (1.47,0.34).

>>>>

Tables \@ref(tab:YTresultsallstrata), \@ref(tab:YTresultsreducedstrata), \@ref(tab:CodERROR), and \@ref(tab:HadERROR) contain the results of our comparison between the absolute error between abundance estimates and model output. General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors and standard deviations. On the contrary, we also see individual cases where the stratified mean produced the lowest absolute error. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

In considering the Yellowtail Flounder results in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata), we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are included on sampling, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \@ref(tab:YTresultsallstrata) and \@ref(tab:YTresultsreducedstrata). These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.

@@

<font size="5">*Differences Between VAST Settings*</font>

For YTF, 29/96 ~ 30% of VAST runs with new settings were better (70% were worse). 80/96 ~ 83% of scenarios had a VAST run with a better error than the stratified mean. 15/16 of the times the stratified mean was better VAST was not using covariates. The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop_IncTemp_Allstrata_WCov_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP_IncTemP_allstrata for Had and YT). The overestimate in the fall only seems to be related to the way the population shifts between season. Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata). In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17). Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates

@@

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low. This can be further examined by viewing estimate/ratio ratio plots, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.

A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 73\% of all VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios with spring VAST ratios producing (1.08,0.12), fall VAST ratios being (1.50,0.38), stratified mean spring ratios resulting in (0.91,0.10), and fall values of (0.84,0.16). The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1. However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates. When including all strata, the stratified mean produced (0.87,0.12), while adding covariates improved the VAST estimate ratio from (1.45,0.17) to (1.13,0.09). With a reduced number of strata, the stratified mean resulted in our worst summary values of (0.59,0.10), and adding covariates to our VAST models decreased the average ratio results from (1.04,0.33 ) to (0.78,0.19).

### 2023-03-15 Manuscript/Draft1.tex
updated manuscript and ratio plots

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error of abundance estimates and model output. General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors. We also see individual cases where the stratified mean produced the lowest absolute error and instances where including covariates in VAST models actually increase the absolute error. When we reduce the number of strata that are included in indexing calculations to simulation species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal.

@@

In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in the Table. These errors are corrected by including covariate information allowing for an improved VAST estimate that are significantly lower than their stratified mean counterparts. Also for Yellowtail, with all strata included adding covariates greatly improved estimates. With strata excluded there was still an improvement with covariates, but much less dramatic than with all strata.

Notes on differences between VAST settings: For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse). 80/96 \textasciitilde{} 83\% of scenarios had a VAST run with a better error than the stratified mean. 15/16 of the times the stratified mean was better VAST was not using covariates. The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop\_IncTemp\_Allstrata\_WCov\_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP\_IncTemP\_allstrata for Had and YT). The overestimate in the fall only seems to be related to the way the population shifts between season. Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata). In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17). Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates

@@

A more careful visual analysis of all error plots for each species reveals that VAST estimates tend to provide error estimates that are too high, while the stratified mean estimates are, on average, too low. Additionally, model/estimate ratio plots for all VAST estimates remain much closer to the desired value of 1 compared to stratified mean estimates, which can range as high as 10 and exhibit large yearly changes.

>>>>

Tables \ref{tab:YTresultsallstrata}, \ref{tab:YTresultsreducedstrata}, \ref{tab:CodERROR}, and \ref{tab:HadERROR} contain the results of our comparison between the absolute error between abundance estimates and model output. General themes are that VAST estimates provide lower errors relative to those derived from the stratified mean, with VAST models that include covariate information providing the lowest overall errors and standard deviations. On the contrary, we also see individual cases where the stratified mean produced the lowest absolute error. When we reduce the number of strata that are included in indexing calculations to simulate species shifting into new territory, we typically see an increase in absolute error (as expected), though there are some scenarios where the impact is minimal. Furthermore, there are scenarios where including covariates in VAST models actually increases the absolute error, especially when we fail to sample the entire domain.

In considering the Yellowtail Flounder results in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}, we can see VAST estimates generally provide lower errors relative to those derived from the stratified mean, with models that include covariate information typically providing the lowest errors. When all strata are included on sampling, adding covariates greatly improved estimates in all scenarios. We still see an improvement in VAST estimates when certain strata are excluded from sampling, but the improvement was much less dramatic than with all strata included. However, there are several instances in which VAST failed to provide improved abundance estimates during the fall season without covariate information, producing the largest errors seen in Tables \ref{tab:YTresultsallstrata} and \ref{tab:YTresultsreducedstrata}. These errors are corrected by including covariate information resulting in improved VAST estimates that are significantly lower than their stratified mean counterparts.

@@

\emph{Differences Between VAST Settings}

For YTF, 29/96 \textasciitilde{} 30\% of VAST runs with new settings were better (70\% were worse). 80/96 \textasciitilde{} 83\% of scenarios had a VAST run with a better error than the stratified mean. 15/16 of the times the stratified mean was better VAST was not using covariates. The 1 time that VAST was using covariates and was still worse than the stratified mean was IncPop\_IncTemp\_Allstrata\_WCov\_WNoise in the Fall (VAST had strong overestimate in the fall during IncPOP\_IncTemP\_allstrata for Had and YT). The overestimate in the fall only seems to be related to the way the population shifts between season. Looking at the percent shift plots, the population is shifting out of strata 16 in both seasons (large east strata). In the spring the population is shifting mostly into strata 17 (thin strata adjacent to 16), but in the fall they are shifting into both 17 and 18 (18 thin one adjacent to 17). Thus it seems like with much of the population concentrated in the really small outer most strata, VAST produces an underestimate, even with covariates

@@

\emph{Model / Estimate Ratio Results}

A simple visual analysis of all error plots for each species reveals that VAST estimates tend to provide abundance estimates that are above the true model value, while the stratified mean estimates are, on average, too low. This can be further examined by viewing estimate/ratio ratio plots, where we divide the yearly abundance estimates by the true model value. In doing so we see that VAST estimates tend to remain closer to the desired value of 1 compared to stratified mean estimates, which can range as low as 0 and exhibit large yearly changes.

A more careful analysis of individual yearly model/estimate ratios when all strata are included revealed that about 73\% of all VAST ratios were above 1 (27\% less than 1), with an average of 1.29 and a standard deviation of 0.21. On the other hand, 33\% of stratified mean estimates were above 1 with an average of 0.874 and a standard deviation of 0.12. There were seasonal differences in estimate ratios with spring VAST ratios producing (1.08,0.12), fall VAST ratios being (1.50,0.38), stratified mean spring ratios resulting in (0.91,0.10), and fall values of (0.84,0.16). The breakdown for individual species followed a similar pattern.

When the entire domain is sampled, adding covariates brings the estimate ratio closer to 1. However, when failing to sample the entire domain, adding covariates can decrease the accuracy of VAST estimates. When including all strata, the stratified mean produced (0.87,0.12), while adding covariates improved the VAST estimate ratio from (1.45,0.17) to (1.13,0.09). With a reduced number of strata, the stratified mean resulted in our worst summary values of (0.59,0.10), and adding covariates to our VAST models decreased the average ratio results from (1.04,0.33 ) to (0.78,0.19).
